---
title: "Pretty Scree Plot - Parallel Analysis and CIs for Eigenvalues "
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ScreePlot_ConfidenceIntervalls_ForObservedEigenvalues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

One of the most pressing and important question
in factor analysis (fa) or principale component analysis (pca)
is the number of factors to retain 

# Example Results Pretty - Scree-Plot + CI 
For PCA & EFA

PICTURE 2B Inserted

# Example Scree-Plot including CIs

For the popular mtcars dataset 

```{r setup}
library(datscience)
library(boot)
# Run a Parallel Analysis (Horn, 1965), save the parallel object
parallel_analysis <- psych::fa.parallel(mtcars, fa="both")
```


## Create Bootstrapped Confidence Intervals for observed Eigenvalues

### PCA Eigenvalues CIs

```{r pcaeigen, warning=FALSE}
PCA_bootObj <- booted_eigenvalues(mtcars, iterations = 500, fa = "pc")
PCA_bootObj

PCA_CIs <- getCIs(PCA_bootObj)
PCA_CIs
```

### EFA Eigenvalues CIs

```{r faeigen, warning=FALSE}
EFA_bootObj <- booted_eigenvalues(mtcars, iterations = 500, fa = "fa")
EFA_bootObj

EFA_CIs <- getCIs(EFA_bootObj)
EFA_bootObj
```

### Create Pretty Scree

```{r prettyscree}
PCA_Plot <- pretty_scree(parallel_analysis,fa="pc")
EFA_Plot <- pretty_scree(parallel_analysis,fa="fa")
```

### Adding Confidence Intervals for Observed Eigenvalues

```{r addCIs}

PCA_Plot <- add_ci_2plot(PCA_Plot,PCA_CIs,type="errorbars")
EFA_Plot <- add_ci_2plot(EFA_Plot,EFA_CIs,type="band")

PCA_Plot
EFA_Plot

```


