---
title: "Pretty Scree Plot with CIs for Eigenvalues "
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ScreePlot_ConfidenceIntervalls_ForObservedEigenvalues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"

  # TODO: Add Citations and References. Elaborate or refer to choice between fa and PCA
)
```
## Purpose

Conducting a factor analysis (fa) or principale component analysis (pca) can be 
challenging. It starts with the choice between either `fa` or `pca`. And continues
with the question on how many factors or components to retain.
In this vignette we will look especially into the second question. How to determine the
number of factors/components to retain with parallel analysis (HORN citation), 
and what this package provides to help with the process.Especially we will look into the possibility to add bootstrapped confidence intervals around the observed Eigenvalues.

#### Recommended Reads

The following are helpful sources (e.g., on the question on how to determine the number of Factors / Components)
_to be advanced, article citations_

* [Tutorial](http://personality-project.org/r/psych/HowTo/factor.pdf) by William Revelle et al. on how to conduct a factor analysis with the psych package.

#### Example Output: A "Pretty" Scree-Plot + added CIs 
_to be inserted picture of the result_

## "Pretty" Scree with CIs a Tutorial

Below you find a step-by-step tutorial in just 4 easy steps on how to create 
bootstrapped confidence intervals (CIs) around the observed Eigenvalues in parallel analysis. 

### 1. Preparation and Example Data
  
This first section will setup the necessary packages and provide a first glance
at the example survey data (on personality) we will need for the tutorial.

##### Loading Packages 

We start by loading all required packages. For this vignette we will use the following packages:

* [`datscience`](https://buedenbender.github.io/datscience/index.html), 
* [`dplyr`](https://dplyr.tidyverse.org/) 
* [`flextable`](https://davidgohel.github.io/flextable/) 
* [`psych`](https://cran.r-project.org/web/packages/psych/psych.pdf)   

I personally like [`pacman`](https://www.rdocumentation.org/packages/pacman/versions/0.5.1) 
package manager for loading libraries in R. But that's up to your preference, you might as well just you `library()` instead

```{r}
pacman::p_load(datscience, psych, dplyr, flextable)
```

##### Example Data

We will use the `bfi` data, that comes with the `psych` package as an example here.
The data set contains information on $k = 25$ items from a Big Five personality questionnaire and
additional $k = 3$ items on sociodemograhpic data from $n = 2800$ participants.
More details can be obtained with `?bfi`.  
  
For this vignette we will just use the 10 items on agreeableness and extraversion,
and sample randomly 300 participants in order to improve the run time.
Below you can get first insights into the data.frame. For the overview we used the [`datscience::format_flextable`](https://buedenbender.github.io/datscience/reference/format_flextable.html) function for the apperance of the table.

```{r}
set.seed(42) # necessary for reproducible results
data(bfi)
# Select only agreeableness and extraversion and sample participants
df <- bfi %>%  
    select(matches("^(E\\d|A\\d)")) %>% 
  # Regex run down 
  #   ^ : starts with
  #   (|) : Logical or
  #   E\\d: E followed by any digit, e.g. E3 matches E\\d
  #   A\\d: A followed by any digit, e.g. A5 matches E\\d
  sample_n(300)
  
dim(df)
str(df)
class(df)
df %>%
  head(5) %>%
  flextable() %>%
  datscience::format_flextable(.,table_caption = c("Table 1", 
                                                 "Glimpse at the First 5 Rows of Shortened bfi Data"))
```

### 2. Running a Parallel Analysis

The psych package makes it pretty convenient to conduct a parallel analysis.
The `fa` parameter allows you to specify if you want to run a principal component
analysis (`fa = "pca"`), or a factor analysis (`fa = "fa"`) or both (`fa = "both"`).
As we want to showcase CI for the observed Eigenvalues for both method we chose
"both" here.


```{r setup}
# Run a Parallel Analysis (Horn, 1965) with psych pacakge and save the parallel object
parallel_analysis <- psych::fa.parallel(df, fa="both")
```
  
  
### 3. Create Bootstrapped Confidence Intervals for observed Eigenvalues

To get bootstrapped Eigenvalues we just use two functions from the `datscience`
package. First we need to create a `boot` object with `booted_eigenvalues()`,
additionally supplying the method (either `pc` or `fa`) and the data.frame.
Afterwards we will extract the confidence interval (CI) with the function 
`getCIs` by passing the boot object.

#### 3.1 Bootstrapped CIs for PCA

```{r pcaeigen, warning=FALSE, results='hide'}
PCA_bootObj <- booted_eigenvalues(df, iterations = 500, fa = "pc")
PCA_bootObj

PCA_CIs <- getCIs(PCA_bootObj)
PCA_CIs
```

### EFA Eigenvalues CIs

```{r faeigen, warning=FALSE, results='hide'}
EFA_bootObj <- booted_eigenvalues(df, iterations = 500, fa = "fa")
EFA_bootObj

EFA_CIs <- getCIs(EFA_bootObj)
EFA_bootObj
```

### 4. Plotting

#### 4.1 Create Pretty Scree Plot

The output of the `psych::fa.parallel` can be easily beautified by the
`pretty_scree` function (below we exemplarily showcase the pretty scree for pca).


```{r prettyscree, warning=FALSE}
PCA_Plot <- pretty_scree(parallel_analysis,fa="pc")
EFA_Plot <- pretty_scree(parallel_analysis,fa="fa")

PCA_Plot
```

### 4.2 Adding Confidence Intervals for Observed Eigenvalues

Adding the obtained CIs to the plot is also pretty easy with the function
`add_ci_2plot()`. You can either chose to display the CIs as bands (with 
`type = "band"` or as error bars with `type = "errorbars"` )

```{r addCIs, warning=FALSE}
EFA_Plot <- add_ci_2plot(EFA_Plot,EFA_CIs,type="band")
PCA_Plot <- add_ci_2plot(PCA_Plot,PCA_CIs,type="errorbars")

EFA_Plot
PCA_Plot
```
