---
title: "A Guide to Bivariate Table 1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flex_table1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Sample Characteristics (Table 1) 
### Including Bivariate Comparisons

Before starting this tutorial, I want to thank the developers of the packages

 * [`table1`](https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html) (Benjamin Rich), and 
 * [`flextable`](https://ardata-fr.github.io/flextable-book/) (David Gohel and Colleagues).
 
Really all credit should go to the teams maintaining these two packages.  
**Full disclosure:** My function merely provides an easy to use API or wrapper around their packages to get
a beautiful publication ready bivariate comparison table 1.

So what am I actually talking about, you might think? 
See for yourself below, it just requires one function call!

```{r eval=FALSE}
# Only 2 mandaotry arguments: formula specifying variables to be used & data (see also table1::table1())
flex_table1(
  str_formula = "~ random_Smoking  + Glucose + Insulin + Leptin +Age + BMI|Diagnosis", 
  data = breast_cancer_modified,
  table_caption = c(
    "Table 1",
    "Sample Characteristics, Comparison of Healthy Controls and Cancer Patients."
  )
```



```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
library(datscience)
# Additionally load Tidyverse for Data Wrangling
library(dplyr)
library(labelled)
library(flextable)

path <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00451/dataR2.csv"
breast_cancer <- read.csv(path)

breast_cancer_modified <- breast_cancer %>%
  mutate(Diagnosis = factor(Classification,
    labels = c("Healthy Controls", "Breast Cancer Patients")
  )) %>%
  select(-Classification) %>%
  mutate(random_Smoking = sample(c("Smoker", "Non-Smoker", "Occasionally"),
    size = nrow(breast_cancer),
    replace = TRUE, prob = c(0.4, 0.4, 0.2)
  )) %>%
  # Convert the Randomly Generated Variable into a Factor
  mutate(random_Smoking = factor(random_Smoking))

var_label(breast_cancer_modified$random_Smoking) <- "Fictional Smoking Status"

ft <- flex_table1(
  str_formula = "~ random_Smoking  + Glucose + Insulin + Leptin +Age + BMI|Diagnosis", data = breast_cancer_modified,
  table_caption = c(
    "Table 1",
    "Sample Characteristics, Comparison of Healthy Controls and Cancer Patients."
  )
)

ft %>%
  set_table_properties(layout = "autofit", width = 1)
```

 In published articles this kind of table (above) is usually 
the first table (thus called Table 1). The example above shows the descriptive stats 
of two subgroups of the sample (aka bivariate descriptive).
The function also automatically conducts group comparisons (and if desired corrects)
the obtained p-values.

 * For **metric / interval** data: independent sample t-test or in case of heterogenity of variances the Welch correction is applied
 * For **categorical / ordinal** data: Pearon's `r knitr::asis_output("\u03C7\u00B2")`-test or respective Fisher\'s test for expected cell counts of `r knitr::asis_output("\u2264")` 5).
 
  I will show you how you can create such a table and get it into word with just a 
few basic and simple steps. 

# Step by Step Tutorial

So how do we get there, just follow these few and easy steps.

## 1.) Data and Setup

### 1.1) Load Required Packages

We will primarly use the `datscience` package. As well as a bit of `dplyr` and `labelled` to
prepare the data. However please note that the function I am showcasing here heavily
benefits from the two package [`table1`](https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html) [`flextable`](https://ardata-fr.github.io/flextable-book/) (as described above).   
If you want to customize the appearance
of your table I recommend you to check out the original sources. The function and vignette
provided here just serve for convenience purposes, to make the creation of a bivariate
comparison in table1 a less time consuming task. 

**Loading Packages**
  
```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(datscience)
# Additionally load dplyr and labelled for Data Wrangling
library(dplyr)
library(labelled)
# We are also setting seed because we will generate random data with sample below
set.seed(123)
```  
  
  
The data we are using for this tutorial is publicly [Patricio, 2018] available on the homepage of the [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Coimbra).
First, we will import the data with `read.csv`, additionally we will slightly modify
the original data set. 

For modification we will start by changing the column `Classification` to a factor and assigning the correct labels, obtained
from the UCI repository.:

```
| Labels:
| 1=Healthy controls
| 2=Patients
```

Additionally we will generate a new fictional variable called `random_Smoking` with 3 associated factor levels `c("Smoker","Non-Smoker","Occasionally")`.   

**Please note** that this is not part of the original data and was just fabricated to illustrate the utility of the `flex_table1()` function on comparison of factors.

**Loading & Modifying the Data**

```{r data}
# Load the data
path <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00451/dataR2.csv"
breast_cancer <- read.csv(path)

# Modify the Data
breast_cancer_modified <- breast_cancer %>%
  # Create a Factor Variable from Classification Column
  # And add Labels to the Factor Levels
  mutate(Diagnosis = factor(Classification,
    labels = c("Healthy Controls", "Breast Cancer Patients")
  )) %>%
  # Remove the old Variable
  select(-Classification) %>%
  # Generate a New Fictional Categorical Variable: random_Smoking
  mutate(random_Smoking = sample(c("Smoker", "Non-Smoker", "Occasionally"),
    size = nrow(breast_cancer),
    replace = TRUE, prob = c(0.4, 0.4, 0.2)
  )) %>%
  # Convert the Randomly Generated Variable into a Factor
  mutate(random_Smoking = factor(random_Smoking))

# Giving the new fictional column a name to be shown in the table
var_label(breast_cancer_modified$random_Smoking) <- "Fictional Smoking Status"
```

Just to get a glimpse and overview of the data.
Additionally, this give the opportunity to showcase another 
useful function of this (`datscience`) package the `format_flextable()`

**Inspection of the Data**

```{r}
# determine how many number of decimal places by convention
nod <- get_number_of_decimals(nrow(breast_cancer_modified))
head(breast_cancer_modified) %>% 
  # round numeric columns
  mutate(across(where(is.numeric), ~ round(.x,nod))) %>% 
  flextable() %>% 
  format_flextable(table_caption = c("Table 2","First Cases (N = 5) of the Cancer Patients Data Set"))


```
Additionally lets inspect distributional parameters and descriptive statistics
with the use of the function `psych::describe()`, and format the output
again with `datscience::format_flextable()`. All outputs generated here can also be
conveniently saved to word with `datscience::save_flextable()`

**Descriptive Statistics and Distribution**

```{r}
breast_cancer_modified %>% 
  # remove categorical variables
  select(where(is.numeric)) %>% 
  # get descriptives of metric variables
    psych::describe() %>%
    as.data.frame() %>% 
    round(nod) %>% 
  # add rownmaes (name of variable) as a separate columns
    tibble::rownames_to_column(., var = "Variable") %>% 
    select(-vars) %>% 
  # generate flextable
    flextable() %>% 
  # format flextable
    format_flextable(table_caption = c("Table 3","Complete Sample - Numeric Variables: Distribution and Descriptive Stat."))
# determine how many number of decimal places by convention
nod <- get_number_of_decimals(nrow(breast_cancer_modified))
head(breast_cancer_modified) %>% 
  # round numeric columns
  mutate(across(where(is.numeric), ~ round(.x,nod))) %>% 
  flextable() %>% 
  format_flextable(table_caption = c("Table 2","First Cases (N = 5) of the Cancer Patients Data Set"))
```



## 2.) Creation of Table 1

The function `datscience::flex_table1` provides a nice API for creating Table 1
with bivariate group comparison. There are only two mandatory arguments:

 * data = the data.frame to use, in the example we use the previously prepared modified breast cancer data `data = breast_cancer_modified`
 * str_formula = the variables that should be displayed and compared in formula notation. _Note_ formula is given as a string in this case. It starts allways with a tilde `"~"` followed by the variables to be shown in the rows combined with a plus sign `"Age + BMI"`. Lastly the groups to be compared a specified after `"| Diagnosis"`.  For more details please refere to the [table1 documentation](https://www.rdocumentation.org/packages/table1/versions/1.4.2/topics/table1).  Lets say for our cancer sample we want to know if there are difference in the following variables
    + Glucose, Insulin, Leptin, Age and BMI of the sample, Additionally if the fictional smoking status differs
    + Between Cancer Patients and Healthy Controls
    + The formula string would look like this: `"~ random_Smoking  + Glucose + Insulin + Leptin +Age + BMI|Diagnosis"`
  
**Creation of the Table**


```{r message=FALSE, warning=FALSE, paged.print=TRUE}
flex_table1(
  str_formula = "~ random_Smoking  + Glucose + Insulin + Leptin +Age + BMI|Diagnosis", 
  data = breast_cancer_modified,
  table_caption = c(
    "Table 4",
    "Sample Characteristics, Comparison of Healthy Controls and Cancer Patients."
  )
)%>%
  # Only For markdown we additionally need to fix the autofit property, this step is not
  # needed when directly saved to word e.g. with datscience::save_flextable()
  set_table_properties(layout = "autofit", width = 1)



```


# Literature
[Patricio, 2018] Patrício, M., Pereira, J., Crisóstomo, J., Matafome, P., Gomes, M., Seiça, R., & Caramelo, F. (2018). Using Resistin, glucose, age and BMI to predict the presence of breast cancer. BMC Cancer, 18(1)


